---
title: "Data Services Landscape Scan"
author: "Data Cooperative"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(kableExtra))
```

# Summary (bulleted)

+ Why did we do this?
+ How did we do this?
+ Brief statement of # institutions and # services
+ Brief statement about results (widespread, rare)
+ UAL compared to peers (above regression line)
+ Next steps? Where should we invest?

# Introduction

Data-intensive scholarship is an area of rapid growth, and many academic 
libraries support this scholarship through data services. The Data Cooperative 
supports several facets of data-intensive scholarship, including data 
management, geospatial data stewardship and analysis, and computational 
literacy. However, there may be opportunities in data services that the 
University of Arizona Libraries have yet to capitalize on.

To identify current UAL strengths and opportunities in supporting data-intensive 
scholarship, we performed a systematic survey of University of Arizona peer 
institutions' library-based data services. Library staff surveyed library 
websites for institutions listed as UA Peers by the University Analytics & 
Institutional Research, including the University of Arizona. The survey was 
designed to determine whether or not a library offered a service by library 
staff, including, but not limited to services in data management, geospatial 
data, statistical consulting, and data visualization. Here we describe the 
survey details and results, discuss the limitations of the approach, and close 
with recommendations for future investments to support data-intensive work at 
the University of Arizona.

# Methods

## Data collection

The landscape scan was designed to determine the prevalence of a variety of 
data services offered at peer institutions of the University of Arizona. The 
25 institutions listed at 
[https://uair.arizona.edu/content/ua-peers](https://uair.arizona.edu/content/ua-peers)
includes ABOR-designated peer institutions as well as all PAC-12 universities. 
We included all 25 institutions in our survey, including the University of 
Arizona. Two UAL employees who were not part of the Office of Digital 
Innovation & Stewardship, performed the first scan of the services offered by 
the libraries at the 25 institutions. The two employees performed this survey 
as part of the 2020 Work From Home effort and were asked to record the URL 
where the service was described using a Qualtrics form. For a subset of the 
services, the modality (web resource, instruction, or consult) was also 
recorded. Only those services that were (1) offered by library personnel (as 
opposed to guest instructors or outside consultants) and in the case of 
instructional sessions or events (2) offered during or after January 2019 were 
considered present at an intitution's library.

Following this initial scan, members of the Data Cooperative first 
checked each URL to ensure the service was described at the URL; this was 
required due to the specialized nature of some services not being correctly 
classified in the initial scan. Each URL indicating the presence of a service 
was independently checked by two members of the Data Cooperative. In some 
cases, this resulted in some services being re-classified as absent. Finally, 
for each service still listed as being absent from an institution, members of 
Data Cooperative again searched the institution library's website to confirm 
that the service was not offered. In some cases, this resulted in some services 
being re-classified as present. 

```{r services-table, echo = FALSE}
# Want to use our dataset to create the table of services we surveyed for; only
# works if each service was found in at least one institution (true as of 
# 2021-08-04)
services <- read.csv(file = "data/services-final.csv")

# Create column that indicates whether or not we surveyed the modality of the 
# service in question
service_types <- services %>% 
  select(Service, Mode) %>% 
  distinct() %>% 
  group_by(Service) %>% 
  dplyr::summarize(NoMode = any(is.na(Mode)))

# Remove the underscores from service names, make some more human accessible 
# and add asterisks to those services that we surveyed for modality
service_table <- service_types %>%
  mutate(Service = gsub(pattern = "_", replacement = " ", x = Service)) %>%
  mutate(Service = if_else(Service == "Geospatial", 
                           true = "Geospatial services (general)", 
                           false = Service)) %>%
  mutate(Service = if_else(Service == "Data management",
                           true = "Data management services (general)",
                           false = Service)) %>%
  mutate(Service = if_else(Service == "OSF",
                           true = "Open Science Framework (OSF)",
                           false = Service)) %>%
  mutate(Service = if_else(NoMode, 
                           true = Service, 
                           false = paste(Service, "*"))) %>%
  select(Service)

# Print out the table
knitr::kable(x = service_table,
             format = "pipe",
             # centering = FALSE,
             caption = "Services included in this landscape scan; asterisks (*) 
             indicate those services for which delivery mode (web resource, 
             instruction, or consultation) was also assessed.")
```


We were also interested in testing how library resourcing might affect the 
number of services offered by an academic library. As a measure of library 
resourcing, we used library salary and wage expenditures for 2019. These data 
were retrieved from the 
[Integrated Postsecondary Education Data System](https://nces.ed.gov/ipeds/use-the-data)
(IPEDS) survey data on 2021-05-17. While this measures _total_ salary and wage 
expenditures for a library, it is the best available proxy for measuring 
resourcing to research-oriented services at academic libraries.

## Analysis

Following quality control checks described above, we measured the frequency of
services offered by institutions' libraries. We counted the total number of 
services offered by each library, regardless of which modalities they were 
offered through. We also counted the number of libraries offering each 
services, to identify which services are common among UAL peers, and which 
services are rare.

We used the total count of services offered at each library to test for an 
influence of library resources. Briefly, we used linear regression to test 
whether the total library salary and wage expenditures predicted the total 
number of services offered. We removed those data points which were categorized 
as having disproportionate influence on regression results. All analyses were 
performed using the [R software package](https://www.R-project.org/).

## Limitations

The systematic web survey method has been used in prior work (e.g. 
[Radecki & Springer 2020](https://doi.org/10.18665/sr.314397)), but it is 
important to note some caveats with the approach. First, surveying web sites 
relies on efficient navigational design - some services may have been 
categorized as absent by our scan if library web sites had poor search 
optimization or non-intuitive menus and navigation. In some cases it was 
difficult to determine whether or not a service listed on a library's web site 
was still being offered. Finally, as this scan took place during the COVID-19 
pandemic, library service offerings were undoubtedly evolving rapidly. However, 
as web sites range from petrifyingly static to changing weekly, we could not 
discern specific impacts of COVID-19 library service offerings.

# Findings

## How many services per library?
## How many libraries for each service?

![**A**. Frequency of services supported across institutions; bars shown in red represent services offered by UAL; grey bars indicate services not supported by UAL. **B**. Number of services across institutions; vertical dotted black line is average number of services offered by libraries (10.64) and red vertical line indicates number of services offered by UAL (12)](output/service-dist-ranks-az.png)

## Services by salary
![Total number of services offered as a function of library's total annual salaries and wages. Dotted line shows regression excluding outlier Washington State University (lower left corner); solid line shows regression excluding Washington State University and Stanford University (upper right corner); red dot (upper left) is UAL.](output/salaries-services-az.png)

# Insights / Conclusions?
Areas where we are like our peers
  + Note ubiquity of data management support
  + Of those offered by over half institutions (n = 9), seven are either data 
  management or geospatial
Are we doing anything special (i.e. the rarities)?
  + OSF
Where could we send more resources?
  + TDM, web scraping ==> Python
  + version control (just more expertise)

# Benefits & outcomes
1 paragraph describing benefits to relevant stakeholders
1 paragraph about measurable outcomes; can be integrated with benefits description

# Closing summary
1 paragraph re-iterating problem and solution
