---
title: "Data Services Landscape Scan"
author: "Data Cooperative"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
---

```{r setup, echo = FALSE}
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(kableExtra))
```

# Summary

+ To guide future directions of data services offered by the University of 
Arizona Libraries, we performed a landscape scan of services offered at the 
libraries of peer institutions.
+ We performed a systematic web survey to determine which of 18 services were 
offered at the libraries of 25 institutions (including the University of 
Arizona).
+ Data management and geospatial data services were common across institutions, 
while support for web scraping and version control were rare.
+ UAL offered an above-average number of these data services.
+ The number of services directly supported by UAL is higher than expected 
given library resourcing.
+ We recommend continued support for data management and increased personnel 
to support geospatial applications and computational humanities and social 
sciences services.

# Introduction

Data-intensive scholarship is an area of rapid growth, and many academic 
libraries support this scholarship through data services. The Data Cooperative 
supports several facets of data-intensive scholarship, including data 
management, geospatial data stewardship and analysis, and computational 
literacy. However, there may be opportunities in data services that the 
University of Arizona Libraries have yet to capitalize on.

To identify current UAL strengths and opportunities in supporting data-intensive 
scholarship, we performed a systematic survey of University of Arizona peer 
institutions' library-based data services. Library staff surveyed library 
websites for institutions listed as UA Peers by the University Analytics & 
Institutional Research, including the University of Arizona. The survey was 
designed to determine whether or not a library offered a service by library 
staff, including, but not limited to services in data management, geospatial 
data, statistical consulting, and data visualization. Here we describe the 
survey details and results, discuss the limitations of the approach, and close 
with recommendations for future investments to support data-intensive work at 
the University of Arizona.

# Methods

## Data collection

The landscape scan was designed to determine the prevalence of a variety of 
data services offered at peer institutions of the University of Arizona. The 
25 institutions listed at 
[https://uair.arizona.edu/content/ua-peers](https://uair.arizona.edu/content/ua-peers)
includes ABOR-designated peer institutions as well as all PAC-12 universities. 
We included all 25 institutions in our survey, including the University of 
Arizona. We used a systematic web survey of the institutions' library web sites 
to assess which services were supported directly by the library. Two UAL 
employees who were not part of the Office of Digital Innovation & Stewardship, 
performed the first scan of the services offered by the libraries at the 25 
institutions. The two employees performed this survey as part of the 2020 Work 
From Home effort and were asked to record the URL where the service was 
described using a Qualtrics form. For a subset of the services, the modality 
(web resource, instruction, or consult) was also recorded. Only those services 
that were (1) offered by library personnel (as opposed to guest instructors or 
outside consultants) and in the case of instructional sessions or events (2) 
offered during or after January 2019 were considered present at an intitution's 
library.

Following this initial scan, members of the Data Cooperative first 
checked each URL to ensure the service was described at the URL; this was 
required due to the specialized nature of some services not being correctly 
classified in the initial scan. Each URL indicating the presence of a service 
was independently checked by two members of the Data Cooperative. In some 
cases, this resulted in some services being re-classified as absent. Finally, 
for each service still listed as being absent from an institution, members of 
Data Cooperative again searched the institution library's website to confirm 
that the service was not offered. In some cases, this resulted in some services 
being re-classified as present. 

```{r services-table, echo = FALSE}
# Want to use our dataset to create the table of services we surveyed for; only
# works if each service was found in at least one institution (true as of 
# 2021-08-04)
services <- read.csv(file = "data/services-final.csv")

# Create column that indicates whether or not we surveyed the modality of the 
# service in question
service_types <- services %>% 
  select(Service, Mode) %>% 
  distinct() %>% 
  group_by(Service) %>% 
  dplyr::summarize(NoMode = any(is.na(Mode)))

# Remove the underscores from service names, make some more human accessible 
# and add asterisks to those services that we surveyed for modality
service_table <- service_types %>%
  mutate(Service = gsub(pattern = "_", replacement = " ", x = Service)) %>%
  mutate(Service = if_else(Service == "Geospatial", 
                           true = "Geospatial services (general)", 
                           false = Service)) %>%
  mutate(Service = if_else(Service == "Data management",
                           true = "Data management services (general)",
                           false = Service)) %>%
  mutate(Service = if_else(Service == "OSF",
                           true = "Open Science Framework (OSF)",
                           false = Service)) %>%
  mutate(Service = if_else(NoMode, 
                           true = Service, 
                           false = paste(Service, "*"))) %>%
  select(Service)

# Print out the table
knitr::kable(x = service_table,
             format = "pipe",
             # centering = FALSE,
             caption = "Services included in this landscape scan; asterisks (*) 
             indicate those services for which delivery mode (web resource, 
             instruction, or consultation) was also assessed.")
```

We were also interested in testing how library resourcing might affect the 
number of services offered by an academic library. As a measure of library 
resourcing, we used library salary and wage expenditures for 2019. These data 
were retrieved from the 
[Integrated Postsecondary Education Data System](https://nces.ed.gov/ipeds/use-the-data)
(IPEDS) survey data on 2021-05-17. While this measures _total_ salary and wage 
expenditures for a library, it is the best available proxy for measuring 
resourcing to research-oriented services at academic libraries.

## Analysis

Following quality control checks described above, we measured the frequency of
services offered by institutions' libraries. We counted the total number of 
services offered by each library, regardless of which modalities they were 
offered through. We also counted the number of libraries offering each 
services, to identify which services are common among UAL peers and which 
services are rare.

We used the total count of services offered at each library to test for an 
influence of library resources. Briefly, we used generalized linear regression 
to test whether the total library salary and wage expenditures predicted the 
total number of services offered. We tested to see if any data points had 
disproportionate influence on regression results, but none were identified as 
having undue influence. All analyses were performed using the 
[R software package](https://www.R-project.org/) and all 
data and analytical code are available on [GitHub](https://github.com/UAL-RE/data-service-landscape-scan).

For a subset of the services, we also analyzed if the general area of service 
(data management, geospatial, or data science) influenced the mode (web 
resource, instruction, or consult) by which the service was offered. We 
analyzed each of the three areas separately with logistic regression to 
determine if there were significant differences in the likelihood of a service 
being offered in each modality. We used Tukey _post-hoc_ pairwise comparisons 
to compare individual modes to one another.

## Limitations

The systematic web survey method has been used in prior work (e.g. 
[Radecki & Springer 2020](https://doi.org/10.18665/sr.314397)), but it is 
important to note some caveats with the approach. First, surveying web sites 
relies on efficient navigational design - some services may have been 
categorized as absent by our scan if library web sites had poor search 
optimization or non-intuitive menus and navigation. In some cases it was 
difficult to determine whether or not a service listed on a library's web site 
was still being offered. Finally, as this scan took place during the COVID-19 
pandemic, library service offerings were undoubtedly evolving rapidly. However, 
as web sites range from petrifyingly static to changing weekly, we could not 
discern specific impacts of COVID-19 library service offerings.

# Findings

Data services offered at peer academic libraries broke into two distinct groups:
those services offered at a clear majority of the libraries surveyed and 
those "rareties" that were only offered at a minority of libraries (Figure 1A). 
Support for data management in general as well as data management plans was 
ubiquitous, found at all surveyed libraries. Geospatial-oriented services, 
including geospatial software and aerial imagery services were also found at a 
majority of the libraries investigated here. At the rare end, support for 
electronic laboratory notebooks and web scraping were found in fewer than 25% 
of libraries we surveyed. Libraries provided support for an average of 10.6 of 
the services, while UAL supports 12 services, by our survey assessment (Figure 
1B, also see [Insights](#insights), below)

![**A**. Frequency of services supported across institutions; bars shown in red 
represent services offered by UAL; grey bars indicate services not supported by 
UAL. **B**. Number of services across institutions; vertical dotted black line 
is average number of services offered by libraries (10.64) and red vertical 
line indicates number of services offered by UAL (12)](output/service-dist-ranks-az.png)

The amount of resourcing, as measured by a libraries' total salary and wage 
expenditures, influenced the total number of data services offered. In a 
regression analysis, salary and wage expenditures was significantly correlated 
with the total number of services a library offered (_p_ = 0.046; line in 
Figure 2).

![Total number of services offered as a function of library's total annual 
salaries and wages. Solid line shows generalized linear regression model; red 
dot (upper left) is UAL.](output/salaries-services-az.png)

The mode in which a service was offered differed among the different areas of 
data services (Figure 3). Data management services (data management support in 
general, and support for data management plans) were more likely to be offered 
through consults and web resources than through formal instructional events (p 
= 0.001 and p = 0.009, respectively). In contrast, data science services (data 
analysis and data visualization software support) were more likely to be 
offered through instructional opportunities than web resources (p = 0.003). 
Geospatial services (general geospatial data support and geospatial software 
support) were equally likely to be offered via any of the three modes (web, 
consult, or instruction).

![Modes of delivery for the different data service areas: data management, 
geospatial support, and data science (data analysis and data visualization 
software.)](output/service-modes.png)

# Insights

The University of Arizona Libraries performed well in providing data-oriented 
services, especially considering the performance of peer institutions. UAL 
supports all of the common data management and geospatial data services (Figure 
1A) and offers an above-average total number of data services (Figure 1B). When 
considering the level of resourcing in terms of salaries and wages, UAL 
provides more services than expected (i.e. it is above the regression line in 
Figure 2). Although UAL was in the lower 25<sup>th</sup> percentile of terms of 
resourcing, the number of services offered was above the average number of 
services offered across surveyed institutions.

The services not directly supported by UAL were generally also not supported at 
peer institutions (Figure 1A). Some of these services, including version 
control, statistical consulting, and web scraping, require specialized 
expertise that may not be common in library personnel. These services 
categorized as "unsupported" at UAL do highlight a limitation of our web site 
survey approach: statistical consulting and support for version control are 
both provided through the R programming sessions, but are not explicitly 
mentioned on UAL web sites. This is largely due to the limited capacity of the 
Data Cooperative: advertising support for these services, especially 
statistical consulting, would be disingenuous, given the high demand for such 
services on campus and limited number of personnel who can support such 
services.

## Recommendations

UAL provides considerable support for data services, primarily through the 
Research Engagement department. Based on our survey findings, there are 
opportunities for deepening engagement and broader support for data-intensive 
work at the University of Arizona.

+ **Continued commitment to data management support**: The recent 
reorganization will allow high-quality support for data management, 
reproducible research, and open science to expand.

+ **Increased support for geospatial data services**: Nearly as ubiquitous as 
data management support, geospatial data services are in increasingly high 
demand as GIS-based applications are used across domains, including public 
health and public policy. To provide a quality service in this area, we 
recommend a 0.5 FTE graduate assistantship to provide operational geospatial 
support. The current model of _ad hoc_ student positions detracts from 
important developmental work in support of geospatial services.

+ **Hire Digital Scholarship faculty with expertise in computational humanities 
& social sciences**: While a minority of libraries provide support for 
computational approaches like text data mining and web scraping, these are 
areas of growing need on many campuses. This includes the University of 
Arizona, as evidenced by the Computational Social Science graduate certificate 
program. Investing in library personnel with expertise relevant to these fields 
would allow UAL to be a leader in providing such services. Furthermore, such 
investment would allow a sharing of the workload necessary to support services 
that overlap with the area of data science, including version control, data 
analysis, and statistical consulting.

# Summary

Support for data services _sensu lato_ at UAL is strong, and future investments 
in this area will further our relationship with researchers and scholars doing 
data-intensive work. Data management support is a stalwart in academic 
libraries, including UAL, and demands for these services are likely to continue 
to increase with the increased emphasis on reproducibility from publishers and 
funding sources. Geospatial and GIS applications are also drawing more 
attention across academic institutions, which contrasts with UAL's decreasing 
staff support in this area - continued quality support will rely on an 
appropriate level of resourcing. Finally, UAL has the opportunity to lead in 
supporting data and computational literacy across campus by meaningfully 
supporting digital scholarship. While adjacencies with data science have been 
sufficient to support topics like programmatic text data mining, intentional 
support through UAL digital scholarship faculty could deepen the relationship 
with colleges such as Social & Behavioral Sciences, whereby UAL partners with 
researchers and scholars in the humanities and social sciences.